Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-10-26 demystify.tex
updates

\subsection{Should you contact a professor? How to get a desired reply?}\label{sec:contact}

@@

\subsection{Kiss of Death in Emails}\label{sec:kiss-of-death-emails}

\begin{itemize}
    \item \textbf{Send emails to the wrong prof.}. While this is not mentioned, it is the number one reason why you don't get a reply: you send email to profs. who do not / cannot advise CS PhD students (e.g., adjunct, emeritus, teaching, non-CS, etc.). See \autoref{sec:faculty-types} for details.
    \item \textbf{Generic}. A copy and paste kind of email or those that can be sent to multiple ppl with very little modifications. This shows the lack of interest and will be treated as spam.  
    \item \textbf{Self-focus}. Focusing too much about you and your achievements but not why you are interested in the prof.'s work (\autoref{sec:contact}). Mention why you're interested in their work and how your background can contribute. 
    \item \textbf{Too long}. Keep it to about 3--4 short paragraphs. Less is more and too long emails are often not read and discarded.\\
    Personally I do not want to see course transcript or test scores in the first email. If I am interested I'll ask for them. Attaching CV is OK. Sample papers and links to your Arxiv papers or GitHub projects are also OK if they are relevant to my work.  
    \item \textbf{Flowery greetings and language}. Don't use ``Dear esteemed professor'', don't ``hey there''.  Do not call the prof. by their first name in the first email (some don't care but you don't want to take the risk -- you don't know them that well yet).  Do not use Mr., Mrs., etc. To be safe, use Prof. Lastname or Dr. Lastname (\autoref{sec:address}).
    \item \textbf{Ignoring the Prof's guideline} and asking questions that are already answered on their website.  Most CS profs have websites and many put very specific information on how to contact them (e.g., email subject, what to include).  Following this helps you stand out and increase your chance of getting a reply.
    \item \textbf{Fancy format}. CS (and probably many other fields) prefers plain email.  Do not use colors, fonts, or fancy formats.
    \item \textbf{Mass emails}. I've seen it many times when a student mass emails all profs in a department (e.g., through CC or even BCC---we know because faculty talk to each other).  This will result in no reply or a very harsh one on how unprofessional you are.
\end{itemize}

>>>>

\section{Contacting a prof. How to get a desired reply?}\label{sec:contact}

@@

\section{Kiss of Death in Emails}\label{sec:kiss-of-death-emails}

\begin{itemize}
    \item \textbf{Send emails to the wrong prof.}. Most students often do not pay attention or know about this, but a very common reason why you don't get a reply is that you send email to profs. who do not / cannot advise CS PhD students (e.g., adjunct, emeritus, teaching, non-CS, etc.). See \autoref{sec:faculty-types} for details. While these profs. might be able to co-advise, they typically do not have the bandwidth, funding, and the desire to mentor students (they are already overloaded with teaching and other service duties). So do some research before sending emails, e.g., most CS profs. who are active in research will have a website with their work (publications) and students; some also have lab website dedicated to their group. Most teaching and adjunct profs. do not have such a research page.

    \item \textbf{Generic}. You should already know this! A copy and paste kind of email or those that can be sent to multiple ppl with very little modifications show the lack of interest and will be treated as spam. Most likely we will not reply to these emails.

    \item \textbf{Self-focus}. Focusing too much about you and your achievements but not why you are interested in the prof.'s work (\autoref{sec:contact}). Mention why you're interested in their work and how your background can contribute.

    \item \textbf{Too long}. Keep it to about 3--4 short paragraphs. Less is more and too long emails are often not read and discarded.\\
    Personally I do not want to see course transcript or test scores in the first email. If I am interested I'll ask for them. Attaching CV is OK. Sample papers and links to your Arxiv papers or GitHub projects are also OK if they are relevant to my work.  
    \item \textbf{Flowery greetings and language}. Don't use ``Dear esteemed professor'', don't ``hey there''.  Do not call the prof. by their first name in the first email (some don't care but you don't want to take the risk -- you don't know them that well yet).  Do not use Mr., Mrs., etc. To be safe, use Prof. Lastname or Dr. Lastname (\autoref{sec:address}).

    \item \textbf{Ignoring the Prof's guideline} and asking questions that are already answered on their website.  Many profs. put very specific information on how to contact them on their website (e.g., email subject, what to include).  Following this helps you stand out and increase your chance of getting a reply.

    \item \textbf{Fancy format}.  Do not use colors, fancy fonts or formats. While not really a kiss of death, it is very annoying, especially for ppl in CS (and probably many other fields) who often prefer plain messages.

    \item \textbf{Mass emails}. I've seen it many times when a student mass emails all profs in a department (e.g., through CC or even BCC---we know because faculty talk to each other).  This will result in no reply or a very harsh one on how unprofessional you are.

    \item \textbf{Don't CALL}. Not related to email but sometimes students get desperate and call the prof.  This is a big no-no, especially for CS ppl who often prefer email over phone calls. Seriously though, don't call any prof. unless they ask you to. 
\end{itemize}

Even if you do everything right, you will likely get a generic (\autoref{sec:interpreting-response}) or no response from a prof. Don't take it personally. They might not have the bandwidth to take on new students (very common), or they might not be interested in your background.

\section{Interpreting Response}\label{sec:interpreting-response}

Some generic responses are:
\begin{itemize}
\item \textbf{No reply}. Very common. You might try again in a few weeks or months, but don't expect a reply. And after a couple of tries, you should move on.  It simply means they are not interested.
\item \textbf{Not taking students but encourage you to apply}. Polite way of saying not interested and referring you to the admission process.
\item \textbf{Not taken student this year (but encourage you to apply next year)}. Polite but generic response.  But If you do apply next year, you can mention that you contacted them last year and they encouraged you to apply.
\item \textbf{Come talk to me after you're admitted}. Generic. Refer you to the admission process. But if you get admitted then you can reply to them and say you're admitted and would like to talk.
\item \textbf{Cannot admit you directly. Need to go through admission process}. Generic. They are not interested and refer you to the admission process.
\item \textbf{I am not taking students but I co-advise/can serve on your committee} While this might sound good, it's generic because it says once you're admitted and have an adviser, then contact me again. 
\item \textbf{I am not taking student but Prof. X might be}. Not common practice as most profs. will not refer you to their colleagues. However, this is better than the previous responses. While they cannot take you, they think you are a good fit for X. So follow up with a thank you and say you'll contact X.  And then contact X and say that Y referred you to them.
\end{itemize}

In short, all of these replies mean the prof. is not interested.  The best positive response is that they want more information from you, e.g., your CV, transcript, paper, or a chat (like an interview).
